Closing the gap between statistical and scientific workflows for improved forecasts in ecology

### 2025-10-18 revision/r2/forecastflows_r2.tex
Small edits to main text

\begin{abstract}
% Concerns about increasing biodiversity loss and climate change have led to greater demands for useful ecological models and forecasts. Datasets relevant for developing these models and forecasts have also increased in size and complexity, including in their geographical, temporal and phylogenetic dimensions. New research often suggests that models accounting for these complexities can yield more accurate trends and predictions. We argue, however, that the usual workflows for model fitting in ecology make it difficult to evaluate and compare most current models for several reasons. First, the research community is split between two disconnected paradigms: either using data to fit simple, trend-like models with one or few parameters, or developing forecasting models that include complex, mechanistic, and/or black box submodels only indirectly informed by data. Second, in both cases, models tend to be developed not only in isolation of one another, but also without a coherent framework for linking scientific questions and understanding to statistical and modeling decisions, throughout the entire modeling process. We propose a unified, principled workflow for end-to-end empirical model development that bridges the gap between process-based and statistical approaches, integrates sound statistical and scientific practices, and especially relies on data simulation to inform decisions at multiple steps in the process. We argue that this approach, coupled with a shift toward universal training, more open model sharing, and alignment on common datasets, could transform for ecological modeling.
Concerns about biodiversity loss and climate change have led to greater demands for ecological models. Datasets for developing these models have increased in size and complexity, including in their geographical, temporal and phylogenetic dimensions. New research often suggests that models accounting for these complexities can yield more accurate trends and predictions. 
We argue, however, that the usual workflows for model fitting in ecology make it difficult to evaluate and compare current models. First, the research community is split between two spheres that prevent uniting a connection between trend estimation and forecasting. One research sphere focuses on using data to fit simple, trend-like models with few parameters, which is separated from research developing forecasting models, often including complex, mechanistic sub-models indirectly informed by data. Second, in both cases, models tend to be developed without a coherent framework for linking scientific questions and understanding to statistical and modeling decisions. To address these challenges we propose a workflow that integrates statistical and scientific practices through clear steps, many of which rely on data simulation to inform decisions in the process, and where forecasting a natural output. We show how this approach, coupled with a shift toward universal training, more open model sharing, and alignment on common datasets, could harmonize currently divided efforts at trend estimation and forecasting to better inform sustainable policies.

@@

Once we are confident about our model structure, we can introduce real data as part of an initial model fitting step (step 4, Figure \ref{fig:workflow}). This way, we obtain parameter estimates constrained by observations (\llabel{quad7}post-model/post-data, Figure \ref{fig:modeldata}d). 
These parameter estimates lead to the second data simulation step, this time using our fitted model parameters to generate predictions (step 5, Figure \ref{fig:workflow}). This---which we call a retrodictive check (\citealp{betanworkflow}, also called a \llabel{PPC}posterior predictive check in a Bayesian workflow,  \citealp{Gelman2020})---allows model output to be compared to observations \llabel{suppfig1}(see Supplementary Figure 1). \llabel{howtoPPC} This is a powerful approach, in part because checks should be tailored to assess key features of the data and model output together in regard to our research question, but this also means there is no standard way to do this \citep{Gelman2020}.
It is only once all steps have been completed that we can interpret parameter values with respect to our research question.
The workflow encourages a focus on the full model, where any parameter (such as a trend estimate) must be carefully interpreted alongside others, as all are fundamental components that shape both inference and forecasting.

@@

We argue that an improved workflow that required data simulation and retrodictive checks would lead to larger model advances and a greater recognition of uncertainty---thus highlighting likely consistency in estimates across models---that could better aid policy.  Using the workflow would make it more obvious that what now appear as major discrepancies are shifts in point estimates that are generally all in the same uncertainty space \citep{Johnson2024}---and it would challenge modelers to show major predictive advances, which is not currently part of the process. Explanatory power in most models of observational data is usually very low \citep{low2014rising,moller2002much} and thus tests of models' predictions rarely expected. But the workflow highlights that predictions from the model---what we call retrodictive checks---are part of the process of science, and critical to testing for what may be missing in a model \llabel{suppfig3}(see Supplementary Figure 1).

We expect retrodictive checks on most published trend analyses would highlight major missing components in these models, and drive changes both in the models themselves and in the simulated data to check the models. This step builds somewhat on the skills needed for null models \citep{Gotelli:2012oz}, but with a shift in focus towards the specific bespoke model at hand. Ecologists have started to use simulated data more to understand potential limitations of their models and data combined \citep{Hilborn1997}, but this is still extremely rare, and efforts to date often treat simulations as separate from the statistical model \citep{Buschke2021,dove2023quantifying}, short-circuiting their full utility if used in an iterative workflow.

@@

\noindent % Process-based models are built on explicit mathematical equations to describe (supposedly causal) relationships between environmental drivers and ecological responses. 
% They also often incorporate empirical relationships, particularly when knowledge is incomplete or when some processes are intentionally omitted. Processes are often represented at different nested spatiotemporal scales, depending on the underlying assumptions. 
Model development is the central step of the process-based workflow, typically requiring several years, yet it often remains opaque for anyone who has not worked the model itself. The step of designing the model---translating knowledge and hypotheses into mathematical equations and parameters---is often blurred with the step of model calibration (or tuning), where parameter values are inferred. Models are often treated as an accumulation of multiple submodels, each governing one or several ecological processes (see Figure B above). Rather than being fitted as a whole, submodels are calibrated separately against specific subsets of data, and some parameters are simply prescribed (i.e., fixed to a value found in the literature) or tuned to reproduce certain observations or theory. The way models are currently calibrated is likely not a coincidence, but rather a workaround to avoid confronting the full complexity of the model. Calibrating submodels separately makes possible to avoid the fact that, if the model were fitted as a whole, many parameters would compensate for one another---revealing structural degeneracies and making the model far more difficult to use.  %emw24Jul: nice stuff here.

@@

% New paragraph by Jim!
\llabel{studydesignandmore}Beyond routine application of the workflow and universal training to support it, the full transformative potential of the workflow hinges on additional broader systemic changes within the research community. First, fostering more open model sharing through standardized repositories and collaborative platforms would accelerate scientific discovery by allowing researchers to build upon, validate, and refine existing models more efficiently, Second, open sharing of more comprehensive study design and sampling design details associated with datasets would enable more appropriate data integration and model design decisions, and independent validation of these decisions by the community. Third, where feasible, alignment on common datasets would provide a standardized basis for model development and comparison, reducing the fragmentation of efforts and facilitating more direct and meaningful evaluations of different modeling approaches. Coupled with universal training, these shifts in the sharing and standardization of data and models would create a more collaborative, efficient, and rigorous environment for ecological modeling, ultimately leading to more accurate forecasts and better understanding of complex environmental systems.

@@

With rapid advances in machine learning, improving current methods to gain greater scientific insights to drive better forecasting seems increasingly important. \llabel{clearerML} A careful and coherent workflow is essential to preserve the interpretability of non–machine-learning models---their primary advantage. \llabel{MLvsPBM}Machine learning will likely surpass process-based models for forecasting accuracy, especially if the latter lack a robust estimation of their parameters and fall in a complexity trap---a trap with additional costs to interpretability of process-based models. Similarly, estimates of trends from empirical data using models without clear mechanistic drivers may soon offer fewer advantages over machine learning.

@@

Today model development in ecology is rarely transparent, which limits how easily the research community can understand models, and thus identify potential issues. Instead of broad inclusive conversations about how to improve models to advance our ecological understanding, a significant portion of scientific debate has become mired in methodological considerations. However, we believe our workflow provides a tractable step to fixing this. By focusing on model development more tightly tied to ecological expertise, we argue this workflow should broaden the community that contributes to model development. It may also help resolve apparent conflicts by identifying where divergences in model predictions emerge, whether from differences in assumptions, model structures, or other aspects of the modeling process.
As ecologists are increasingly expanding their computational toolkits, many field, and lab, and other forms of `empirical' ecologists have the basic tools to follow this workflow to build models that better represent their ecological domains of interest, and---most importantly---to interrogate those models.

>>>>

\begin{abstract}
% Concerns about increasing biodiversity loss and climate change have led to greater demands for useful ecological models and forecasts. Datasets relevant for developing these models and forecasts have also increased in size and complexity, including in their geographical, temporal and phylogenetic dimensions. New research often suggests that models accounting for these complexities can yield more accurate trends and predictions. We argue, however, that the usual workflows for model fitting in ecology make it difficult to evaluate and compare most current models for several reasons. First, the research community is split between two disconnected paradigms: either using data to fit simple, trend-like models with one or few parameters, or developing forecasting models that include complex, mechanistic, and/or black box submodels only indirectly informed by data. Second, in both cases, models tend to be developed not only in isolation of one another, but also without a coherent framework for linking scientific questions and understanding to statistical and modeling decisions, throughout the entire modeling process. We propose a unified, principled workflow for end-to-end empirical model development that bridges the gap between process-based and statistical approaches, integrates sound statistical and scientific practices, and especially relies on data simulation to inform decisions at multiple steps in the process. We argue that this approach, coupled with a shift toward universal training, more open model sharing, and alignment on common datasets, could transform for ecological modeling.
Concerns about biodiversity loss and climate change have led to greater demands for ecological models. Datasets for developing these models have increased in size and complexity, including in their geographical, temporal and phylogenetic dimensions. New research often suggests that models accounting for these complexities can yield more accurate trends and predictions. 
We argue, however, that the usual workflows for model fitting in ecology make it difficult to evaluate and compare current models. First, the research community is split between two spheres that prevent uniting trend estimation and forecasting. One research sphere focuses on using data to fit simple, trend-like models with few parameters, which is separated from research developing forecasting models, often including complex, mechanistic sub-models indirectly informed by data. Second, in both cases, models tend to be developed without a coherent framework for linking scientific questions and understanding to statistical and modeling decisions. To address these challenges we propose a workflow that integrates statistical and scientific practices through clear steps, many of which rely on data simulation to inform decisions in the process, and where forecasting a natural output. We show how this approach, coupled with a shift toward universal training, more open model sharing, and alignment on common datasets, could harmonize currently divided efforts at trend estimation and forecasting to better inform sustainable policies.

@@

Once we are confident about our model structure, we can introduce real data as part of an initial model fitting step (step 4, Figure \ref{fig:workflow}). This way, we obtain parameter estimates constrained by observations (\llabel{quad7}post-model/post-data, Figure \ref{fig:modeldata}d). 
These parameter estimates lead to the second data simulation step, this time using our fitted model parameters to generate predictions (step 5, Figure \ref{fig:workflow}). This---which we call a retrodictive check (\citealp{betanworkflow}, also called a \llabel{PPC}posterior predictive check in a Bayesian workflow,  \citealp{Gelman2020})---allows model output to be compared to observations \llabel{suppfig1}(see Supplementary Figure 1). \llabel{howtoPPC} This is a powerful approach, in part because checks should be tailored to assess key features of the data and model output together in regard to our research question, but this also means there is no standard approach  \citep{Gelman2020}.
It is only once all steps have been completed that we can interpret parameter values with respect to our research question.
The workflow encourages a focus on the full model, where any parameter (such as a trend estimate) must be carefully interpreted alongside others, as all are fundamental components that shape both inference and forecasting.

@@

We argue that an improved workflow that required data simulation and retrodictive checks would lead to larger model advances and a greater recognition of uncertainty---thus highlighting likely consistency in estimates across models---that could better aid policy.  Using the workflow would make it more obvious that what now appear as major discrepancies are shifts in point estimates that are generally all in the same uncertainty space \citep{Johnson2024}---and would challenge modelers to show major predictive advances, which is not currently part of the process. Explanatory power in most models of observational data is usually very low \citep{low2014rising,moller2002much} and thus tests of models' predictions rarely expected. But the workflow highlights that predictions from the model---what we call retrodictive checks---are part of the process of science, and critical to testing for what may be missing in a model \llabel{suppfig3}(see Supplementary Figure 1).

We expect retrodictive checks on most published trend analyses would highlight major missing components in these models, and drive changes both in the models themselves and in the simulated data to check the models. This step builds somewhat on the skills needed for null models \citep{Gotelli:2012oz}, but with a shift in focus towards the specific bespoke model at hand. Ecologists have started to use simulated data more to understand potential limitations of their models and data combined \citep{Hilborn1997}, but this is still extremely rare, and efforts to date often treat simulations as separate from the statistical model \citep{Buschke2021,dove2023quantifying}, short-circuiting their full utility when used in an iterative workflow.

@@

\noindent % Process-based models are built on explicit mathematical equations to describe (supposedly causal) relationships between environmental drivers and ecological responses. 
% They also often incorporate empirical relationships, particularly when knowledge is incomplete or when some processes are intentionally omitted. Processes are often represented at different nested spatiotemporal scales, depending on the underlying assumptions. 
Model development is the central step of the process-based workflow, typically requiring several years, yet it often remains opaque for anyone who has not worked the model itself. The step of designing the model---translating knowledge and hypotheses into mathematical equations and parameters---is often blurred with the step of model calibration (or tuning), where parameter values are inferred. Models are often treated as an accumulation of multiple submodels, each governing one or several ecological processes (see Figure B above). Rather than being fitted as a whole, submodels are calibrated separately against specific subsets of data, and some parameters are simply prescribed (i.e., fixed to a value found in the literature) or tuned to reproduce certain observations or theory. 
% emwOct18: I think the issue in the last sentence is partly the passive voice, the negation (easier to avoid = less likely), and the changing of verb tenses, so I switched all this. You can switch back if you want, the sentence was not wrong, it was just hard to get through.
The way models are currently calibrated is likely not a coincidence, but rather a workaround to avoid confronting the full complexity of the model. By calibrating submodels separately researchers are less likely to confront structural degeneracies in models (often because many parameters can compensate for others); in contrast, fitting submodels together (as a whole) would reveal such issues at a point when they can be addressed.  %emw24Jul: nice stuff here.

@@

% New paragraph by Jim!
\llabel{studydesignandmore}Beyond routine application of the workflow and universal training to support it, the full transformative potential of the workflow hinges on additional broader systemic changes within the research community. First, fostering more open model sharing through standardized repositories and collaborative platforms would accelerate scientific discovery by allowing researchers to build upon, validate, and refine existing models more efficiently. Second, open sharing of more comprehensive study and sampling design details associated with datasets would enable more appropriate data integration and model design decisions, and independent validation of these decisions by the community. Third, where feasible, alignment on common datasets would provide a standardized basis for model development and comparison, reducing the fragmentation of efforts and facilitating more direct and meaningful evaluations of different modeling approaches. Coupled with universal training, these shifts in the sharing and standardization of data and models would create a more collaborative, efficient, and rigorous environment for ecological modeling, ultimately leading to more accurate forecasts and better understanding of complex environmental systems.

@@

With rapid advances in machine learning, improving current methods to gain greater scientific insights to drive better forecasting seems increasingly important. \llabel{clearerML} A careful and coherent workflow is essential to preserve the interpretability of non–machine-learning models---their primary advantage. \llabel{MLvsPBM}Machine learning will likely surpass process-based models for forecasting accuracy, especially if the latter lack robust estimation of their parameters and fall into a complexity trap---a trap with additional costs to interpretability of process-based models. Similarly, estimates of trends from empirical data using models without clear mechanistic drivers may soon offer fewer advantages over machine learning.

@@

Today model development in ecology is rarely transparent, which limits how easily the research community can understand models, and thus identify potential issues. Instead of broad inclusive conversations about how to improve models to advance our ecological understanding, a significant portion of scientific debate has become mired in methodological considerations. However, we believe our workflow provides a tractable step to fixing this. By focusing on model development more tightly tied to ecological expertise, we argue this workflow would broaden the community that contributes to model development. It may also help resolve apparent conflicts by identifying where divergences in model predictions emerge, whether from differences in assumptions, model structures, or other aspects of the modeling process.
As ecologists are increasingly expanding their computational toolkits, many field, and lab, and other forms of `empirical' ecologists have the basic tools to follow this workflow to build models that better represent their ecological domains of interest, and---most importantly---to interrogate those models.
